\thetitle

### 2013-09-29 tfpie/response.txt

RESPONSE: Clarified.

p12 It's an interesting comment about retaining female students, but
it's not clear how the design of the course might be relevant to
this. Do you have any suggestions why? Could it just be a random
fluctuation?

RESPONSE: We clarified the anecdotal nature of this.

----------------------- REVIEW 3 ---------------------

(1) in the introduction, page 1, the bullet-list enumerates
'significant transitions happen in concert'. In the paper we learn how
to overcome bullet 1, but not bullets 2, 3, 4, and 5.

RESPONSE: Our approach presents OO programming in a context that
avoids _all 5_ of these significant transitions until after the
fundamental concepts of OO programming are taught, as we emphasize.

On page 8, first
paragraph: "In particular: objects, classes, fields and methods,
dynamic dispatch, inheritance, and overriding." are enumerated as
being essential concepts of object-oriented programming that students
have had experience with in the new set-up. The paper explains this
for the first half ('objects, classes, fields and methods') but not
for the second half ('dynamic dispatch, inheritance, and overriding').

@@

Suggestion for improvement: extend section 3 with material how dynamic
dispatch, inheritance, and overriding is presented to students.

RESPONSE: We discuss dynamic dispatch extensively in the paper.  This
is the entire point of section 3.2.  We added a paragraph discussion
inheritance and overriding.

@@

Suggestion for improvement: shorten section 5.1. The extra space can
be used to elaborate on issue (1) above. I don't think this section
requires sub sections, so all text can be organized into a single
section.

RESPONSE: We feel that this discussion is needed.

(3) In the introduction, it is said that 'an alternative approach to
the second semester [that] overcomes these problems and provides a
gradual introduction to OO programming'. However, acting as the
devil's advocate, I can argue that this is not true: halfway semester
2 there is still the abrupt switch to both another language, Java,
*and* another tool set, Eclipse. Stated bluntly, the switch is only
postponed instead of solved.

RESPONSE: We clarify how the transition is less troublesome here.  We
also describe future work that would further alleviate these
difficulties.

### 2013-09-29 tfpie/tfpie.tex
review 3

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
the students' first year.  The most common approach to achieve this
aim is to teach an industrial language in the first course, often
Java, starting on the first day.  Other curricula take a more indirect
route by teaching functional programming in the first semester,
followed by a second semester in Java.  The latter approach is an
improvement over the first, as pointed out by numerous
observers~\cite{dvanhorn:Felleisen2004Structure, local:spolsky}, but
both suffer serious flaws.

@@

\paragraph{Outline} The remainder of this paper is organized as follows:
in section~\ref{sec:background} provides background on \emph{How to
  Design Programs} and the context and constraints involved in the
first year at Northeastern.  Section~\ref{sec:shift} describes our
approach to the second semester, which starts with a small shift in
perspective to bridge the gap between functional programming and
object oriented programming.  Section~\ref{sec:industrial} describes
the path from our pedagogical languages to an industrial object-oriented
programming language.  Section~\ref{sec:related-work} discusses the
relation to existing work and section~\ref{sec:conclusion} concludes.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets and one for landed rockets (an example is given in
appendix~\ref{sec:appendix}).  While in the functional approach it
is easy to use the state-pattern for the \emph{data} representing a
rocket, it is more difficult to have states of \emph{behavior}.  The
typical solution adds conditionals to all of the event handlers.  In
the object-oriented approach, states of behavior are just as natural
as data.  It is therefore straightforward to design programs with easy
to observe invariants such as ``a landed rocket never changes
position.''  In the functional approach, even such simple properties
are more involved to establish.

@@

All but the first two of these changes are unrelated to the
fundamental lessons we hope to teach---the rest merely present
additional hurdles for students.

@@

;; A World is one of 
;; - (new upworld Number)
;; - (new downworld Number)
(define-class upworld
  (fields n)

@@

  ;; on-tick : -> World
  (define (on-tick)
    (cond [(zero? (this . n))
           (new upworld (this . n))]
          [else
           (new world (sub1 (this . n)))]))

>>>>

Many universities and colleges aim to teach their students proficiency
in an industrial object-oriented programming language by the end of
the students' first year.  The most common approach to achieve this
aim is to teach an industrial language in the first course, often
Java, starting on the first day.  Other curricula take a more indirect
route by teaching functional programming in the first semester,
followed by a second semester in Java.  The latter approach is an
improvement over the first, as pointed out by numerous
observers~\cite{dvanhorn:Felleisen2004Structure, local:spolsky,
  dvanhorn:Chakravarty2004Risks, dvanhorn:Ragde2008Chilling}, but
both suffer serious flaws.

@@

\paragraph{Outline} The remainder of this paper is organized as follows:
 section~\ref{sec:background} provides background on \emph{How to
  Design Programs} and the context and constraints involved in the
first year at Northeastern.  Section~\ref{sec:shift} describes our
approach to the second semester, which starts with a small shift in
perspective to bridge the gap between functional programming and
object oriented programming.  Section~\ref{sec:industrial} describes
the path from our pedagogical languages to an industrial object-oriented
programming language.  Section~\ref{sec:related-work} discusses the
relation to existing work and section~\ref{sec:conclusion} concludes.

@@

The move to object-oriented style immediately and naturally leads to
designs that are enabled by organizing programs around objects.  For
example, the state pattern~\cite{samth:GOF} becomes useful almost
immediately.  The programs in figure~\ref{fig:world} animate a rocket
(rendered as a circle in this example) taking off.  An illustrative
follow-up exercise is to animate a rocket that \emph{lands}.  The
natural design is to have two variants for states of the rocket: one
for descending rockets and one for landed rockets (an example is given
in appendix~\ref{sec:appendix}).  While in the functional approach it
is easy to use the state-pattern for the \emph{data} representing a
rocket, it is more difficult to have states of \emph{behavior}.  The
typical solution adds conditionals to all of the event handlers.  In
the object-oriented approach, states of behavior are just as natural
as data.  It is therefore straightforward to design programs with easy
to observe invariants such as ``a landed rocket never changes
position.''  In the functional approach, even such simple properties
are more involved to establish, because all event handlers must be
inspected.

This approach also leads naturally to discussion of inheritance.
Often programs with multiple states wish to share the implementation
of some methods.  We first show that this can be accomplished at the
cost of minor boilerplate with delegation, and then show how
inheritance allows the programmer to avoid duplication and boilerplate
entirely.  Once inheritance is able to group identical methods,
overriding is a natural next step when some but not all of the
implementations are identical across the variants.

@@

All but the first two of these changes are unrelated to the
fundamental lessons we hope to teach---the rest merely present
additional hurdles for students.  However, at this point in the
semester, the students are far better equipped to meet these
challenges.  They are already familiar with objects, classes, and the
other concepts we have covered.  They are also fully engaged in the
class, instead of making the transition in the midst of the transition
between semesters.  Finally, they have now been programming for 50\%
longer than they had at the start of the semester.

@@

;; A World is one of 
;; - (new landed-world)
;; - (new downworld Number)
(define-class landed-world

  ;; to-draw : -> Image
  (define (to-draw) 
    (place-image 
     (circle 10 "solid" "red")
     390 200 (empty-scene 400 400))))

  ;; on-tick : -> World
  (define (on-tick)
    (cond [(zero? (this . n))
           (new landed-world)]
          [else
           (new world (sub1 (this . n)))]))

@@

  ;; on-key : KeyEvent -> World
  (define (on-key k) (new world 400)))
